---
title: "Imputing Missing Values"
date: '2025-10-01'
engine: knitr
type: slides
categories:
- slides
- Week04
format:
  revealjs:
    transition: slide
    background-transition: fade
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
---


```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(naniar)
library(mice)
library(purrr)
library(NHANES)
```

```{r create-data, echo = FALSE}
set.seed(202509)
df <- tibble(
  id    = 1:20,
  age   = sample(c(18:70, NA), 20, replace = TRUE),
  income = sample(c(10^5*(3:6), NA), 20, replace = TRUE),
  score  = sample(c(1:5, NA, NA), 20, replace = TRUE),
  group  = sample(c("A", "B", NA), 20, replace = TRUE)
)
```

# How do we 'fix' datasets with missing values?

# Approach (1): Complete Case Analysis

- Remove samples with missing values: `na.omit`, `na_drop` 

- Use with extreme caution! This is not recommended!!

```{r}
NHANES |> drop_na() |> dim()
```

- None of the NHANES data is complete.

- Generally, this is a very naive strategy 


# Approach (2): Impute with mean

Impute missing values with mean - Again, not recommended!

```{r}
# Mean imputation (bad practice!)
df_mean_imp <- df |>
  mutate(income_imp = if_else(is.na(income), mean(income, na.rm = TRUE), income)) 
```

```{r}
df_mean_imp |> ggplot(aes(x = income_imp)) + geom_histogram()
```
# Impact of Imputing with Mean

Mean is unaffected

```{r}
mean(df_mean_imp$income_imp)
mean(df_mean_imp$income, na.rm = TRUE)
```

SD is lowered - affects confidence intervals of estimates

```{r}
sd(df_mean_imp$income_imp)
sd(df_mean_imp$income, na.rm = TRUE)
```
# Example: Airquality

The `airquality` data is a(n ancient) data set in R capturing daily ozone, solar, wind and temperature measurements for 5 months   


Some of the ozone measurements are missing

```{r}
summary(airquality$Ozone)
```

How should we impute the Ozone values?

# Airquality overview


```{r}
library(GGally)
ggpairs(airquality)
```
# Model Ozone

```{r}
ozone_lm <- lm(Ozone~Temp + Wind, data = airquality)
summary(ozone_lm)
```

# Predict Ozone

Use a linear model to predict Ozone values based on covariates

```{r}
airquality_imp <- airquality |> mutate(
  Ozone_pred = predict(ozone_lm, newdata = airquality),
  imputed = is.na(Ozone),
  Ozone_imp = ifelse(is.na(Ozone), Ozone_pred, Ozone)
)
```


# Predict Ozone


```{r}
ggpairs(data = airquality_imp |> select(-Ozone_pred, -Ozone), 
        mapping = aes(colour = imputed)) 
```


# Idea for Predictive Mean Matching



# Multiple Imputation with `mice`

```{r mice}
# Run multiple imputation
imp <- mice(df, m = 5, method = "pmm", seed = 202509)

# Completed datasets
completed_list <- complete(imp, "all") |> map(as_tibble)

# Example: regression model
models <- map(completed_list, ~ lm(income ~ age + group, data = .x))

# Pool results
pooled <- pool(models)
summary(pooled)
```

# Sensitivity Analysis

```{r sensitivity}
# Complete case mean
cc_mean <- df |> drop_na(income) |> summarise(mean_income = mean(income))


```




# Exercises

  1. Explore missingness patterns in the `NHANES` dataset. Which variables are most affected?
  2. Compare results of a regression model (`BMI ~ Age + Gender`) under:
  - complete-case analysis,
- mean imputation,
- multiple imputation.
3. Discuss: What assumptions are you making under each approach?


# Conclusion

  - Missingness can bias estimates if ignored.
- Mechanisms (MCAR/MAR/MNAR) determine what methods are valid.
- Naïve fixes are often misleading.
- Multiple imputation and likelihood-based methods are modern, principled approaches.
