---
title: "Deterministic Record Linkage"
date: '2025-10-23'
engine: knitr
type: slides
categories:
- slides
- Week09
format:
  revealjs:
    transition: slide
    background-transition: fade
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
execute:
  error: true
  eval: true
  echo: false
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(stringdist)
library(Lahman)
data("LahmanData")
```

# Intro to Statistical Record Linkage

**Record linkage**: identify records that refer to the same entity
 across different datasets 
 
 
**Deterministic linkage**: Exact matches on one or more key fields (e.g., SSN) 

**Probabilistic linkage**: Uses statistical models and similarity scores  

# Examples

- Linking records from hospital and insurance for the same patient  
- Match voter registrations across states  
- Combine data from different sources where names are noisy or incomplete  
 

# Deterministic Linkage
 
If entities are recorded with the same IDs (e.g. country codes, license plates, SSNs) we can join by those values.

Normalized data bases often include IDs

Use joins (`join_left`, `join_inner`, `join_full`) to combine these data sets

# Your Turn {background-color="#006666"}

The package `Lahman` is an R data package consisting of `r nrow(LahmanData)`
data sets detailing different aspects of baseball.


- Identify the players inducted into the Hall of Fame in 2025 by name

- How often was each of the inductees on the ballot before? 

#

```{r}
inductees25 <- HallOfFame |> filter(yearID==2025, inducted=="Y", category == "Player") |>
  left_join(People |> select(nameFirst, nameLast, playerID))
inductees25

HallOfFame |> filter(playerID %in% inductees25$playerID) |> group_by(playerID) |>
  count() |> arrange(desc(n))

```

# Fuzzy Matching

approximate matching of strings $s$ and $t$

Different algorithms based around type of errors: 
e.g. Jaro-Winkler, Jaccard (common elements), Levenshtein (insertions, deletions),
Damerau–Levenshtein (also transpositions), Hammond (needs strings of equal length)

choose based on which errors most likely occur


# Jaro-Winkler String similarity

For two (character) strings $s_1$ and $s_2$ the *Jaro* similarity is defined as 

$$
J(s_1,s_2) = \frac{1}{3} \left( \frac{m}{|s_1|} + \frac{m}{|s_2|} + \frac{m-t}{m}\right),
$$
if there are at least $m > 0$ matching characters between $s_1$ and $s_2$;

$|s_i|$ is the number of the characters in string $s_i$.

$t$ is the number of transpositions (characters that are at most $\left\lfloor {\frac {\max(|s_{1}|,|s_{2}|)}{2}}\right\rfloor -1$ positions apart)

# Winkler adjustment

Emphasize beginning of words

$$
JW(s_1, s_2) = J(s_1, s_2) + \ell p \left (1 - J(s_1, s_2) \right)
$$

- $\ell$ is length of common prefix at the start (up to 4 characters)

- $p$ scaling factor for upwards adjustment of score for common prefixes (should not exceed $1/\ell$); usually set to $p=0.1$. 

Note that JW distance is not a proper distance (no triangle property)


# Jaro-Winkler Properties

$JW(s,t) \in [0,1]$; larger values indicate higher similarity

Implemented in package `stringdist` as `stringdist` with `method = "jw"`


```{r echo=TRUE}
1 - stringdist::stringdist("Halo", "Hello", method = "jw")
1 - stringdist::stringdist("frist", "first", method = "jw")
```


# First Example

```{r}

# Dataset A: census type data
df_a <- tibble(
  id_a = 1:6,
  first = c("John", "Katherine", "Robert", "Maria", "William", "Alicia"),
  last  = c("Smith", "Jones", "Brown", "Lopez", "Chen", "Mendez"),
  birth_year = c(1985, 1978, 1990, 1988, 1983, 1991),
  city = c("Lincoln", "Omaha", "Grand Island", "Kearney", "Norfolk", "Lincoln")
)

# Dataset B: voter registration data with typos and variant spellings
df_b <- tibble(
  id_b = 101:108,
  first = c("Jon", "Kathy", "Rob", "Maria", "Will", "Alisha", "Laura", "Robert"),
  last  = c("Smyth", "Jones", "Brown", "Lopes", "Chen", "Mendez", "King", "Brown"),
  birth_year = c(1985, 1978, 1990, 1988, 1983, 1991, 1992, 1990),
  city = c("Lincon", "Omaha", "Grand Island", "Kearny", "Norfolk", "Lincoln", "Omaha", "Grand Island")
)
```

:::: {.columns}

::: {.column width="48%"}
Census type data

```{r class.output='small'}
#| code-block-border-left: "#31BAE9"
df_a
```
:::
::: {.column width="4%"}
:::
::: {.column width="48%"}
Voter registration type data

```{r class.output='small'}
df_b
```
:::

:::::


# Create Pairs of Records for each Variable

```{r echo=TRUE}
pairs_first <- data.frame(expand.grid(first_a = unique(df_a$first), first_b = unique(df_b$first)))
pairs_last <- data.frame(expand.grid(last_a = unique(df_a$last), last_b = unique(df_b$last)))
pairs_city <- data.frame(expand.grid(city_a = unique(df_a$city), city_b = unique(df_b$city), stringsAsFactors = FALSE))
pairs_years <- data.frame(expand.grid(birth_year_a = unique(df_a$birth_year), birth_year_b = unique(df_b$birth_year)))
```





# Calculate distance between pairs

Jaro-Winkler in package `stringdist` as `stringdist` with `method = "jw"`


```{r echo=TRUE}
pairs_first <- pairs_first |> mutate(
  sim = 1 - stringdist(first_a, first_b, method = "jw")
)
pairs_first |> arrange(desc(sim))
```

# Introduce Thresholds

```{r echo=TRUE}
pairs_last <- pairs_last |> mutate(
  sim = 1 - stringdist(last_a, last_b, method = "jw")
)

pairs_city <- pairs_city |> mutate(
  sim = 1 - stringdist(city_a, city_b, method = "jw")
)
```

and introduce thresholds:

```{r echo=TRUE}
pairs_first <- pairs_first |> mutate(match = sim > 0.6)
pairs_last <- pairs_last |> mutate(match = sim > 0.8)
pairs_city <- pairs_city |> mutate(match = sim > 0.9)
```

# Example: apply thresholds to combined data

Stratify on birth year:

```{r echo = TRUE}
df_ab <- df_a |> full_join(df_b, by="birth_year", suffix = c("_a", "_b"))
df_ab <- df_ab |> 
  left_join(pairs_first |> select(-sim, match_first = match))  |>
  left_join(pairs_last |> select(-sim, match_last = match))  |>
  left_join(pairs_city |> select(-sim, match_city = match))  |>
  rowwise() |>
  mutate(
    overall = sum(match_first + match_last + match_city)
  )
```
#

```{r}
df_ab
```

