---
title: "Probabilistic Record Linkage"
date: '2025-10-23'
engine: knitr
type: slides
categories:
- slides
- Week09
format:
  revealjs:
    transition: slide
    background-transition: fade
    navigation-mode: vertical
    logo: ../../N.svg
    includes:
      in_header: ../../header.html
execute:
  error: true
  eval: true
  echo: false
  message: false
  warning: false
---

```{r setup}
library(tidyverse)
library(stringdist)
```


```{r}

# Dataset A: census type data
df_a <- tibble(
  id_a = 1:6,
  first = c("John", "Katherine", "Robert", "Maria", "William", "Alicia"),
  last  = c("Smith", "Jones", "Brown", "Lopez", "Chen", "Mendez"),
  birth_year = c(1985, 1978, 1990, 1988, 1983, 1991),
  city = c("Lincoln", "Omaha", "Grand Island", "Kearney", "Norfolk", "Lincoln")
)

# Dataset B: voter registration data with typos and variant spellings
df_b <- tibble(
  id_b = 101:108,
  first = c("Jon", "Kathy", "Rob", "Maria", "Will", "Alisha", "Laura", "Robert"),
  last  = c("Smyth", "Jones", "Brown", "Lopes", "Chen", "Mendez", "King", "Brown"),
  birth_year = c(1985, 1978, 1990, 1988, 1983, 1991, 1992, 1990),
  city = c("Lincon", "Omaha", "Grand Island", "Kearny", "Norfolk", "Lincoln", "Omaha", "Grand Island")
)
```

# Probabilistic Matching

Two main issues:

1. Inconsistencies e.g., "Katherine Smith" vs. "K. Smith" vs "Kate Smith"

2. Errors and missing data e.g., typos, formatting differences, or incomplete fields.


Goal: 

For each pair of records, determine: (1) **Match** (same entity), (2) **Do not match**, or (3) **uncertain**


# Fellegi and Sunter

1. Comparison of fields
2. Model match probabilities
3. Use Thresholds to determine matches


# Step 1. Comparison of fields

For each pair of records, we compare corresponding fields (e.g., name, date of birth, ZIP code) and create **comparison vectors**:

$$
\gamma = (\gamma_1, \gamma_2, ..., \gamma_k)
$$

Each $\gamma_i$ measures the level of agreement for field $i$.

| Field | Record A | Record B | Comparison |
|--------|-----------|-----------|-------------|
| First name | "Jon" | "John" | Partial |
| Last name | "Doe" | "Doe" | Exact |
| Birth year | 1982 | 2010 | Disagree |


#  Step 2. Model match probabilities


Each pair of records belongs to one of two latent groups:
$M$: the **match set**, $U$: the **non-match set**

$$
P(\gamma) = P(\gamma \mid M) P(M) + P(\gamma \mid U) P(U)
$$

where

- $P(\gamma \mid M)$: probability of observing agreements among *true matches*  
- $P(\gamma \mid U)$: probability of observing agreements among *non-matches*  
- $P(M)$ and $P(U) = 1 - P(M)$: prior probabilities of match / non-match.


# EM Algorithm

True match status is unobserved; use Expectation–Maximization (EM) algorithm:

1. **E-step:**  
   Estimate match memberships $P(M \mid \gamma)$ for each pair given current parameters.

2. **M-step:**  
   Update $m_j = P(\gamma_j \mid M), u_j = P(\gamma_j \mid U), P(M)$ based on expected match memberships.

Iterate until convergence.

# Step 3: Posterior Probabilities and Thresholding

Posterior match probabilities for each pair:

$$
P(M \mid \gamma) = \frac{P(\gamma \mid M) P(M)}{P(\gamma \mid M) P(M) + P(\gamma \mid U) P(U)}
$$

Use decision threshold to classify matches, e.g. any records with $P(\gamma \mid M)$  above $T_U$ are matches, any records below $T_L$ are non-matches.

Quantify uncertainty: e.g. trigger manual review for pairs that have posterior probabilities in $[T_L, Y_U]$.

# Example with `fastLink`

The `fastLink` package output provides **posterior probabilities** for all record pairs.  


```{r echo=TRUE}
library(fastLink)

fl_out <- fastLink(
  dfA = df_a,
  dfB = df_b,
  varnames = c("first", "last", "birth_year", "city"),
  stringdist.match = c("first", "last", "city"),
  numeric.match = c("birth_year"),
  partial.match = c("first", "last", "city","birth_year")
)
```
#

```{r}
matches <- getMatches(df_b, df_a, fl_out)
matches %>%
  arrange(desc(posterior)) 
```
